Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
Moved proposed method subsection below svd
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@852 ccdd05d5-56c1-461a-bf98-85f6876523b4
Fixed broken equation reference
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@853 ccdd05d5-56c1-461a-bf98-85f6876523b4
some words
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@854 ccdd05d5-56c1-461a-bf98-85f6876523b4

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the filters of the previous section can be approximated to high
accuracy by the expansion in the singular value basis
%
\begin{equation}
y_{ik}(\tau) \approx \sum_j^{\numsvdtmps} \sigma_{jk} v_{ijk} u_{j}(\tau)
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_{jk}$ is the $j^{\mathrm{th}}$ singular value for the
$k^{\mathrm{th}}$ time slice, $v_{ijk}$ is an orthogonal matrix for the
$k^{\mathrm{th}}$ time slice  and $u_{jk}$ is a new orthogonal basis filter set
for the $k^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the
time-slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time_slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute-force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.

>>>>

\subsection{The \lloid\ method}

In the remainder of this section we explore a method for reducing the
computational cost of a time domain search for compact binary coalescence.  We
will show that it is possible to have a theoretically zero latency algorithm
that competes with the computational cost of the standard frequency domain
matched filtering, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\ involves two transformations to the
template waveforms that produce a set of orthogonal filters with far fewer
sample points than the original template waveforms.  The reduction in the
number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples
overall and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate.  This amounts to more than a factor of 100 in the floating point
operations per second required.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points that need to be filtered by another factor
of $\sim 100$.  The combined methods reduce the number of floating point
operations to the level where they are competitive with the conventional
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the filters of the previous section can be approximated to high
accuracy by the expansion in the singular value basis
%
\begin{equation}
y_{ik}(\tau) \approx \sum_j^{\numsvdtmps} \sigma_{jk} v_{ijk} u_{j}(\tau)
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_{jk}$ is the $j^{\mathrm{th}}$ singular value for the
$k^{\mathrm{th}}$ time slice, $v_{ijk}$ is an orthogonal matrix for the
$k^{\mathrm{th}}$ time slice  and $u_{jk}$ is a new orthogonal basis filter set
for the $k^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the
time-slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time-slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute-force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.
